---
title: "Introduction to MicroDemix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MicroDemix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicroDemix)
```

# Introduction

This vignette demonstrates how to use the `MicroDemix` package, a method that decomposes the stool microbiome at compositional level to understand the heterogeneity of the gut microbiome across various GI locations.

## Load the following R packages

```{r}
library(optimx)
library(cubature)
library(rootSolve)
library(MASS)
library(ggplot2)
library(BB)
library(MCMCpack)
```

## Example datasets in 'MicroDemix'

The example datasets are from the simulation study with 10 taxa and 100 samples. 

### 'yig': A stool (mixture) microbiome dataset

'yig' is an example dataset for stool (mixture) microbiome data with 100 samples in rows and 10 taxa in columns. We can access this dataset by running the following:

```{r}
data(yig)
head(yig)
```

### 'yig.n': A rectum (reference) microbiome dataset
'yig.n' is an example dataset for rectum (reference) microbiome data with 100 samples in rows and 10 taxa in columns. We can access this dataset by running the following:

```{r}
data(yig.n)
head(yig.n)
```

### 'x': A covariates dataset related to samples in data 'yig'
'x' is a dataset contains covariates corresponding to samples in data 'yig'. We can access this dataset by running the following:

```{r}
data(x)
dim(x)
```

### 'x.n': A covariates dataset related to samples in data 'yig.n'
'x.n' is a dataset contains covariates corresponding to samples in data 'yig.n'. We can access this dataset by running the following:

```{r}
data(x.n)
dim(x.n)
```

## Functions and data analysis with 'MicroDemix'

### Function 1: MD_estimate(data1, data2, cova)

This function estimates the microbial relative abundance in stool (mixture), rectum (reference) and other GI locations. For example:

```{r}
est1 <- MD_estimate(yig, yig.n, x)
est1
```

### Function 2: RA_plot(p.r, p.s, p.o, G, taxon)

We can visualize microbial relative abundances (proportions) at three different locations using the 'RA_plot()' function. The following example replicates one of the plots from the paper, analyzing the phylum Firmicutes at the family level.

```{r, fig.width=7.5, fig.height=4.5}
p.r <- c(0, 0, 0.0195415822, 0.52043881, 0.0002028398, 0, 0.3991845842, 0, 0.0606321839)
p.s <- c(0.02713150778, 0.00075219743, 0.00073529412, 0.43284820825, 0.00010311021, 0.00008958756, 0.53113421231, 0.00650608519, 0.00069979716)
p.o <- c(0.0571234686474, 0.0014037570334, 0.0000033854869, 0.122693737055, 0.0000030544905, 0.0000021475416, 0.8050093482991, 0.0137601784493, 0.0000009229968)
families <- c("Clostridiaceae", "Enterococcaceae",    "Erysipelotrichaceae", "Lachnospiraceae", "Peptococcaceae",  "Planococcaceae",  "Ruminococcaceae",    "Turicibacteraceae", "Veillonellaceae")
RA_plot(p.r, p.s, p.o, 9, families)
```

### Function 3: MD_pvalue(p.r, p.o, N.r, N)

We obtain a simulation-based p-value for testing differential abundance under the null hypothesis that p.r = p.o, where p.r and p.o are the microbial relative abundance in the rectum and other GI locations, respectively.

```{r}
MD_pvalue(est1$pr.est, est1$po.est, sum(yig.n), sum(yig))
```

We can obtain a p-value under other circumstances, for example:

```{r}
p1 <- rep(0.1, 10)
p2 <- c(rep(0.09, 5), rep(0.11, 5))
MD_pvalue(p1, p2, 5000, 5000)
```

### Function 4: (data1, data2, cova, maxiter=10)

This function using the EM algorithm to estimate the microbial relative abundance in stool (mixture), rectum (reference) and other GI locations. The EM algorithm is more computationally intensive due to its relatively slow convergence rate and the repeated use of the Metropolis-Hastings (MH) algorithm in the E-step. In the following example, each EM iteration takes approximately 2 minutes on a local machine. However, the estimation accuracy improves significantly after just a few iterations. 

```{r, eval=FALSE}
system.time(est2 <- MicroDemix_EM(yig, yig.n, x, maxiter=5))
#> [1] "EM iteration: 1"
#> iter:  0  f-value:  4704.532  pgrad:  281.5245 
#> [1] 0.002658093
#> [1] "EM iteration: 2"
#> iter:  0  f-value:  4701.743  pgrad:  277.8827 
#> [1] 0.0008061163
#> [1] "EM iteration: 3"
#> iter:  0  f-value:  4701.414  pgrad:  276.3458 
#> [1] 0.0002402273
#> [1] "EM iteration: 4"
#> iter:  0  f-value:  4701.446  pgrad:  276.3473 
#> [1] 0.0004213928
#> [1] "EM iteration: 5"
#> iter:  0  f-value:  4701.244  pgrad:  274.4289 
#> [1] 0.0001306647
#>    user  system elapsed 
#>  531.33    0.98  940.91

est2
#> $pr.est
#>     Taxon1     Taxon2     Taxon3     Taxon4     Taxon5     Taxon6     Taxon7 
#> 0.03272567 0.04125933 0.05265133 0.06706233 0.08539033 0.10519300 0.13534633 
#>     Taxon8     Taxon9    Taxon10 
#> 0.17478833 0.22439333 0.08119000 
#> 
#> $ps.est
#>     Taxon1     Taxon2     Taxon3     Taxon4     Taxon5     Taxon6     Taxon7 
#> 0.03213167 0.04096100 0.05220233 0.06671833 0.08465967 0.10574300 0.13572900 
#>     Taxon8     Taxon9    Taxon10 
#> 0.17467567 0.22533033 0.08184900 
#> 
#> $po.est
#>     Taxon1     Taxon2     Taxon3     Taxon4     Taxon5     Taxon6     Taxon7 
#> 0.03191630 0.04084895 0.05203789 0.06658997 0.08438937 0.10594348 0.13587368 
#>     Taxon8     Taxon9    Taxon10 
#> 0.17463619 0.22567804 0.08208611
```

### Function 5: MicroDemix_EM_permutation(data1, data2, cova1, cova2, maxiter=10, B)

This function performs a permutation based test of differential abundance under the null hypothesis that p.r = p.o. The permutation test requires multiple runs of the EM algorithm. Therefore, we recommend performing this procedure using parallel computing or a high-performance machine.

```{r, eval=FALSE}
MicroDemix_EM_permutation(yig, yig.n, x, x.n, maxiter=2, B=10)

# [1] "EM iteration: 1"
# iter:  0  f-value:  4704.429  pgrad:  281.1108 
# [1] 0.002657345
# [1] "EM iteration: 2"
# iter:  0  f-value:  4701.638  pgrad:  277.4302 
# [1] 0.0008550884
# [1] "permutation iter: 1"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4688.334  pgrad:  279.2411 
# [1] 0.0008837178
# [1] "EM iteration: 2"
# iter:  0  f-value:  4688.151  pgrad:  279.6902 
# [1] 0.0003699496
# [1] "permutation iter: 2"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4680.651  pgrad:  280.0751 
# [1] 0.0006625398
# [1] "EM iteration: 2"
# iter:  0  f-value:  4680.5  pgrad:  279.7522 
# [1] 0.0002500549
# [1] "permutation iter: 3"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4704.84  pgrad:  280.6383 
# [1] 0.0008771262
# [1] "EM iteration: 2"
# iter:  0  f-value:  4704.655  pgrad:  280.7079 
# [1] 0.0003814814
# [1] "permutation iter: 4"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4695.358  pgrad:  280.5077 
# [1] 0.001261839
# [1] "EM iteration: 2"
# iter:  0  f-value:  4694.589  pgrad:  280.0775 
# [1] 0.0004829247
# [1] "permutation iter: 5"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4714.963  pgrad:  280.0942 
# [1] 0.0006955507
# [1] "EM iteration: 2"
# iter:  0  f-value:  4714.625  pgrad:  278.8059 
# [1] 0.0002040188
# [1] "permutation iter: 6"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4692.693  pgrad:  279.8199 
# [1] 0.0007145038
# [1] "EM iteration: 2"
# iter:  0  f-value:  4692.438  pgrad:  279.3542 
# [1] 0.0004163259
# [1] "permutation iter: 7"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4701.11  pgrad:  281.4275 
# [1] 0.001500689
# [1] "EM iteration: 2"
# iter:  0  f-value:  4699.981  pgrad:  278.8331 
# [1] 0.0002896197
# [1] "permutation iter: 8"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4688.652  pgrad:  278.8074 
# [1] 0.001133446
# [1] "EM iteration: 2"
# iter:  0  f-value:  4688.138  pgrad:  280.7433 
# [1] 0.0005703671
# [1] "permutation iter: 9"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4708.529  pgrad:  280.8886 
# [1] 0.001269645
# [1] "EM iteration: 2"
# iter:  0  f-value:  4707.73  pgrad:  278.8487 
# [1] 0.0002213754
# [1] "permutation iter: 10"
# [1] "EM iteration: 1"
# iter:  0  f-value:  4712.357  pgrad:  280.8193 
# [1] 0.0005794702
# [1] "EM iteration: 2"
# iter:  0  f-value:  4712.134  pgrad:  279.7022 
# [1] 0.0001730983
# [1] 0  #pvalue
```





